\chapter{Cosmos}

To understand the origin of understanding, we must first understand the origin of chaos.

% some theme about unprovable truths

This story begins a long time ago, at the very beginning of everything.
I do not know how long ago this was, for time itself had not yet gained the meaning it has now.
For time to have meaning, there must be change: change in position, in state, in form: for things which are to stop, and things which are not to begin.
But at the beginning, nothing was not: everything was in a state of undifferentiated unity.

An endless soup: matter and energy in a formless gloop, swirling, coalescing, inter-exchanging and transforming from one to the other.
A cauldron of stuff, extending across the entirety of the baby \foreignlanguage{greek}{κόσμος}; a fine balance of creation and destruction so violently dynamic that its only prevailing feature is its absence of features.
This formless state is the \foreignlanguage{greek}{χάος}.
Though without form, it is not nothing: it is everything.
Everything that will ever be, could ever be (or \textit{would} ever be, as we shall come to see), was in this \foreignlanguage{greek}{χάος}.
I shall refer to this collection of matter, energy, space and time, being ``everything that is the case'', as \textit{the universe}.

What came before the universe would be of great importance if it were not for the fact that it had not a single ounce of meaning.
Time may feel to you like an ever-moving arrow, pointing straight from past to future; nothing can speed it up, slow it down, or change its direction, and so there is always a \textit{before} and an \textit{after}.
The unfolding of events lets us track the movement of this arrow of time: something happens, entropy increases, and time moves forward.
But at the beginning, when the entropy was at its lowest, there is nowhere left to follow.

At the North Pole, every arrow points South.
At the beginning of time, amongst the undifferentiated \foreignlanguage{greek}{χάος}, every arrow points to the future, and so it is towards the future we must go.

Over time, as structure began to crystallise from the formless soup, patterns began to emerge.
Regions of higher and lower density formed, and the higher-density regions began to attract more matter, more energy, and thus grew denser still.

Over time, a hierarchy of structures formed.
First, great filaments of matter, forming a cosmic web, with vast voids in between.
These are the back-bone of the universe, a scaffolding upon which all else shall be built.

These filaments coalesced into clusters of matter, which in turn split up into galaxies.
Spirals, ellipses or irregular shapes emerged, each galaxy a swirling dance of gas, bound together by gravity.

Within galaxies, regions of gas and dust collapsed to form stars.
The stars were born, lived out their lives, and died in spectacular displays of explosive energy.
These explosions seeded the universe with heavy matter, which coalesced into new stars, which themselves lived and died.

Some of these stars were surrounded by excess gas, and from this gas, planets might have been born.
Spheres of rock, metal or gas, orbiting their stars in an endless dance.
Initially, all of these planets were hostile places, bombarded by radiation from their stars, or crushed by their own gravity, or submerged in oceans of acid.

% talk more about explosions, collisions, make everything sound violent

For a long time, the universe continued to evolve in this way: this state of affairs persisted for billions of years; interactions between particles on the smallest possible scale have been a constant throughout the history of the universe, but they have rarely amounted to more than that.
Random collisions, making and breaking of bonds, and the occasional formation of more complex structures, such as rocks, crystals, and simple molecules.

Stars were born and died, galaxies collided, planets formed and were destroyed, and nothing more complex than that really seemed to happen---nothing had \textit{purpose}, because to demand purpose implies a certain degree of self-righteousness; a belief that things should be a certain way, fulfil a certain aim, as though to please whoever or whatever it is that is projecting this purpose onto the universe.
To begin with, things just \textit{were}, and they were beautiful.
Certainly, nothing that could observe, or reflect upon, or understand, any of these events was in existence.

Nothing except us, that is.
For a long time, we were the only observers.
The sole awareness in any otherwise silent cosmos.
Things were simpler back then, because there were no expectations, preconceptions, or assumptions to get in the way of seeing things as they really were.
We were simply part of the universe, experiencing it as it unfolded around us.

But one day, on one rocky planet in a quiet corner of a fairly nondescript galaxy, something changed.